Your time series is (probably) too short

### 2017-06-28 writing/manuscript_draft.Rmd
added to discussion and created cover letter

Number of words: X,XXX 
\vspace{3 mm}

To be submitted to: \emph{XXXXX} (Article)
\vspace{3 mm}

Keywords:

@@

@@

## Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. However, census data is often expensive, requiring a lot of people-hours and equipment. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, I explore two approaches to address this question: one simulation-based and an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. <!--Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses.--> Using a simulation approach, I demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher variability in population size. In addition, I examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. I show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--Here, I stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

## Introduction - need to clean up and reduce text

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010]. Longterm data is necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations). However, there are of course notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually necessary? This question has important scientific and management implications. Scientifically, we need to know how long a time series is required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations). From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not. For example, the IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].

In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power, effect size, alpha, sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$), or false negative. In other words, it is a measure of how well you could detect a trend given it is actually significant. Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, as data points are clearly non-independent.

<!--For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987]. Simple population models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker). In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series (cite). Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.-->

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized it for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance. However, these studies, and other past work (citation), have focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. In the main manuscript, we determine the minimum time series length to determine if a population is incresing or decreasing according to a significant slope coefficient in simple linear regression. 
<!-- In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.  -->

## Methods

### Theoretical approach

@@

In Figure \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$). It is clear that statistical power increases quickly with increases in length of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).

```{r,echo=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power.\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

  years_to_sample=seq(2,100,by=1)
  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300 trials

@@

  # Plotting commands
  par(mfrow=c(2,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)

    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=min_time_needed,lty=2,lwd=2,col='red')
      abline(h=0.8,lty=2,lwd=2,col='red')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)

    # Plot versus trend strength
      r_values = seq(1,3,by=0.2)
      sigma_values=seq(1,7,by=0.2)

@@

      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)

@@

      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
```

```{r, echo=F}
# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
  setwd("~/Desktop/Research/time-series-project")

@@

#Load our cleaned data (see other code for cleaning steps)
  load('cleaned-data/pop_info_example_goodNAs.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
  pop_info=pop_info[pop_info$trend_p_value<0.05,]
  long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
  long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
```

### Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]`), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
```
For a subset of populations in the database (X = XX), we were able to connect a database on biological characteristics like body size and generation length. We used

### Empirical approach

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, they are a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). A power of 0.8 would imply that given a trend exists, there is a 0.8 probability of detecting the trend. Complementary, there would be a 0.2 probability of a false negative.

```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of ANIMAL NAME over time. The red line is the best fit line from linear regresssion. (b) statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
# Example of subsampling empirical data
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/pop_info_example_goodNAs.Rdata")
  par(mfrow=c(1,2))
    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='population size',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,at = 1965,font = 2,cex=1.2)
# could insert all possible regressions of length n
source('scripts/calculate_power_metric.R')
    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,at = 7,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)

@@

## Results
<!--
- summary of findings
- theoretical approach
  - example plot
- empirical approach
  - plots: distribution of min time, what min time correlations with
  - walk through one empirical example (maybe in appendix)
-->

@@

### Simulation approach

@@

### Empirical test

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/pop_info_example_goodNAs.Rdata")

@@

```{r,echo=F,fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}

# Plot minimum time required against trend strength, population variance, and autocorrelation  
  par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  mtext('| trend strength |',1,line=3,cex=1.2)
  mtext('(a)',3,line=-1.5,at=0.0318,cex=1.2,font = 2)
    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
    mtext('population variance',1,line=3,cex=1.3)
    mtext('(b)',3,line=-1.5,at=0.128,cex=1.2,font = 2)
      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
      mtext('autocorrelation',1,line=3,cex=1.2)
       mtext('(c)',3,line=-1.5,at=0.97,cex=1.2,font = 2)
       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}). Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Fig. SX). Combined trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.

We also examined a subset of populations where more explanatory variables were known (e.g. body mass, longevity, location). None (need to do regression) of these explanatory variables were predictive of the minimum time series length required (supp mat). Some differences if $T_{min}$ did occur between biological class (supp mat), but this can be attributed to the amount populations are known to fluctuate (i.e. the variability in population size).

```{r,echo=F,include=FALSE, eval=FALSE,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) Longterm trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:biological_correlates}'}
# Biological correlates plots
par(mfrow=c(2,2),mar=c(4,5,0.5,0.5))

boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))

boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
  mtext('| trend strength |',2,line=3.5,cex=1.2)

boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))

boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=1)
stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
```

@@

<!--
## Potential figures
- Examples of how determining minimum time series works (probably in methods)
- What min time should scale with theoretically (trend, power, variability, etc) - another multi panel figure
- Mean min time by class or system
- Separate results for species, body size, location, biology. - multi panel figure

@@

### Questions in the appendix

@@

## Discussion

@@

We show two different approaches to estimate the minimum time series length required required to address a particular question of interest. The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987]. Our simulations support past work...

Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.

First we asked, what is $T_{min}$ for determining whether a population was increasing or decreasing over time according to linear regression? We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig. \ref{fig:fig:min_time_dist}). In line with theoretical predictions (citations), we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (cite figure).

>>>>

Number of words: X,XXX (less than 4,000 words or 20 pages)
\vspace{3 mm}

To be submitted to: \emph{Biological Conservation} (Short Communication)
\vspace{3 mm}

Available as a preprint at: www.peerj.com
\vspace{3 mm}

Keywords: ecological time series, experimental design, monitoring, power analysis, statistical power, sampling design

@@

\pagebreak

@@

# Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. However, census data is often expensive, requiring a lot of people-hours and equipment. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, we explore two approaches to address this question: through simulations and through an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

# Introduction

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010]. Longterm data is necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations). There are, of course, notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others (citations).

Naturally, one might ask, how long of a time series is actually necessary? This question has important implications for both research and management. Scientifically, we need to know the time series length required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations). From a management perspective, it is important to understand when a trend in abundance over time is actually meaningful or not. The IUCN Red List Categories and Criteria suggest under Criterian A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we do not want to sample for longer than is neccessary [@Gerber1999].

In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power (1-$\beta$), effect size, $\alpha$, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$), or false negative. In other words, it is a measure of how well you could detect a trend given it is actually significant. Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987]. Simple population models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker). In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series (cite). Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized it for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance. However, these studies, and other past work, have focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression.
<!-- In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.  -->

# Methods

## Simulation approach

@@

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time series are shown for a set number of time periods ($t=40$). It is clear that statistical power increases quickly with increases in length of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$).

```{r,echo=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

  years_to_sample=seq(2,100,by=1)
  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300+ trials

@@

  # Plotting commands
  par(mfrow=c(2,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=min_time_needed,lty=2,lwd=2,col='red')
      abline(h=0.8,lty=2,lwd=2,col='red')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

    # Plot versus trend strength
      r_values = seq(1,3,by=1) #by=0.2
      sigma_values=seq(1,7,by=1) #by=0.2

@@

      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(c)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
```

<!-- code to pull in databases-->
```{r, echo=F}
# Pull in original database from Keith et al. 2015 and the cleaned version after our series of filtering steps
  setwd("~/Desktop/Research/time-series-project")

@@

#Load our cleaned data (see other code for cleaning steps)
  load('cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
 # pop_info=pop_info[pop_info$trend_p_value<0.05,]
 # long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
 # long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)
```

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics (**cite database here**) like body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.

## Empirical approach

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). A power of 0.8 would imply that given a trend exists, there is a 0.8 probability of detecting the trend. Complementary, there would be a 0.2 probability of a false negative.

<!-- example of how subsampling and empirical approach works -->
```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Harp Seals over time. The line is the best fit line from linear regresssion. (b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
# Example of subsampling empirical data
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n
source('scripts/calculate_power_metric.R')
    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)

@@

# Results
<!--
- summary of findings
- theoretical approach
  - example plot
- empirical approach
  - plots: distribution of min time, what min time correlations with
  - walk through one empirical example (maybe in appendix)
-->

@@

## Simulation approach

@@

## Empirical test

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}'}

# Plot minimum time required against trend strength, population variance, and autocorrelation  
  par(mfrow=c(3,1),mar=c(4,1,1,1),oma=c(0,4.5,0,0.5))
  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  mtext('| trend strength |',1,line=3,cex=1.2)
  mtext('(a)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
    mtext('population variance',1,line=3,cex=1.3)
    mtext('(b)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
      mtext('autocorrelation',1,line=3,cex=1.2)
       mtext('(c)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as expected based on theoretical results (cite papers), the minimum time series length required is strongly correlated with trend strength, population variance, and autocorrelation (Fig \ref{fig:correlates}). Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Fig. SX). Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)` of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.

We also examined a subset of populations where more explanatory variables were known (e.g. body mass, longevity, location). None of these explanatory variables were predictive of the minimum time series length required (**supp mat**). Some differences if $T_{min}$ did occur between biological class (**supp mat**), but this can be attributed to the amount populations are known to fluctuate (i.e. the variability in population size).

@@

<!--
# Potential figures
- Examples of how determining minimum time series works (probably in methods)
- What min time should scale with theoretically (trend, power, variability, etc) - another multi panel figure
- Mean min time by class or system
- Separate results for species, body size, location, biology. - multi panel figure

@@

## Questions in the appendix

@@

# Discussion

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine longterm changes in abundance using linear regression? The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].
